HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-09-22 docs/HLL_exp_bad/main.tex
working through the paper

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Mathilde Raynal, Kenneth G. Paterson\\
\vspace{3mm}
Department of Computer Science, ETH Zurich\\
{\tt mraynal@student.ethz.ch, kenny.paterson@inf.ethz.ch}}

@@

\section{Introduction}
Naive approaches to computing the cardinality of a big data set, such as sorting the elements or simply maintaining the set of unique elements seen, are impractical at scale. The HyperLogLog algorithm (HLL henceforth)~\cite{hll} is today the most widely used cardinality estimator. It is increasingly used in settings where adversaries may have  incentives to manipulate the estimate made by the algorithm. For example,~\cite{portscanhll} proposes its use for detecting network scanning attacks, while Facebook engineers have reported~\cite{fbhll} that they use HLL for estimating the number of active users. These applications demand a careful evaluation of the performance of HLL under adversarial input.

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of $O(2^t)$ where $t$ is the number of bits of flexibility in the input).

The rest of the paper is organised as follows. Section~\ref{sec:overview} gives an overview of the HyperLogLog algorithm. A brief summary of the related works can be found in Section \ref{sec:related}. Section~\ref{sec:attacks} presents our different adversarial models. There, we also give attacks and evaluate their impacts under the different adversarial models. The experimental results are discussed in Section~\ref{sec:exp} and Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

@@

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own register, that we call a bucket. The collection of buckets is referred to as the HLL \emph{sketch}. To map a value $x$ to one of $2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$. The output length of $h(\cdot)$ is thus $n+\ell$, fixed to 32 in~\cite{hll}. Estimators from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ is empirically computed to correct a systematic multiplicative bias.

Relatively small and big cardinalities are handled separately to avoid eventual distortions, as shown in Fig.~\ref{fig:hll}. When the raw cardinality estimate is below $\frac{5}{2}\times 2^n$, the Hit Counting algorithm of Whang et al.~\cite{hitcounting} is used to produce the final cardinality estimator, looking at the number of empty buckets (whose register is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hashing collisions. Some works building on top of~\cite{hll} replace the Hit counting algorithm by other schemes and use a hash function with a larger output domain to reduce the chances of collision and thus get rid of the need for large cardinality range correction. For example, \cite{hllpratice} uses a 64-bit hash function and Linear Counting along with bias correction in the small cardinality range. In this report we target the original paper's implementation reported in Fig.~\ref{fig:hll} and leave for future work the analysis of its variations.

As result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{hll}.

@@

\begin{figure}[ht]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : D \rightarrow \{0, 1\}^{32}$ hash data from $\mathcal{D}$ to binary $32$–bit words.\\
Let $\rho(s)$ be the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^K) = K + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} $\mathcal{M}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $v \in \mathcal{M}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $x := h(v);$\\
    \hspace{2em} \textbf{set} $j = 1 + <x_1x_2...x_{n}>_2$; \hfill \textit{\{the binary address determined by the first $n$ bits of $x$\}}\\
    \hspace{2em} \textbf{set} $w := x_{n+1}x_{n+2}...$;\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the “raw” HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:= m $ln$(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range—no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure}

@@

Regardless of the sensitivity of the above examples, none of the relevant papers are clear about their threat model, especially for the HLL component.  They are all potentially vulnerable to the attacks we present below.

@@

The work of~\cite{cardestprivacy} studies the privacy properties of HLL in two different attack scenarios, referred to as \emph{insider} and \emph{external}. The attack target of~\cite{cardestprivacy} is different from ours, being concerned with loss of privacy, whereas we care about cardinality estimation. It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm. In Section~\ref{sec:attacks} we will discuss the two attack scenarios from~\cite{cardestprivacy} in more detail. In particular, we will show that an attacker who can insert items into the HLL and who can access the HLL cardinality estimate (via typical APIs in HLL implementations) as permitted in the external scenario can actually realise the same attacks as the insider attacker can, at least insofar as cardinality estimate manipulation is concerned.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. Their attack model can be considered to be quite artificial, since it gives the attacker the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. After some clarifications obtained from personal correspondence with the authors, we finally agree that such setup might be possible under certain circumstances. See further discussion in Section~\ref{sec:attacks}.

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as replacement (but at the cost of losing mergeability) or in addition to an unsalted sketch (inducing a memory overhead).
%This might not be a convenient solution for some applications, they thus encourage further research.

@@

\subsection{Adversarial Setting}\label{sec:set}
We model four distinct adversarial scenarios based on the adversary's knowledge and capabilities. To simulate real-life settings, we consider both cases when the sketch is shared and receives inputs from other honest users but also when inputs are under adversarial control only.
It is safe to assume that the adversary knows the setting being used. 
Those two settings differ mainly in the fact that the adversary is attacking a sketch that already contains other users' data in the former option, but is empty in the latter one.

Among the scenarios, the \textit{insider} setup of~\cite{cardestprivacy} that we present as S4 makes the strongest assumptions about the adversarial capabilities since it assumes that the attacker has a perfect view on the sketch.

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a copy of the HLL instance as a user. He can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The attacker does not know the state of the targeted HLL sketch, meaning that he does not have the list of elements previously inserted into it by other users.
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. An S4 adversary can compute a lower bound on the HLL cardinality estimate based on its snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

We start by highlighting the fact that we do not consider scenario S1 typical of HLL deployments. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the attacker learns which machine is used to do the monitoring, buys the same, and uses it as the shadow device. It is possible that the provider uses a different salt or key in the hash per device. This would make any pre-computation on the items using the shadow device irrelevant, and considerably weaken the attack presented in Section 3.2 of~\cite{hllvuln}. Hence, we assume for the rest of this work that shadow and targeted devices have the same internals and security parameters. If the HLL device receives inputs from several users, the attack from Section 3.2 of~\cite{hllvuln} is the easiest to carry out. It works as follow: "after each insertion [in the shadow device], elements that do not increase the cardinality estimate are retained". Iterating through the process several times refines the items' filtering before inserting them into the targeted device. We explore in this work the setup where inputs are coming only from the adversary, and propose an attack that provides better results than the one we just presented.

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the attacker gradually inserts items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once an increase occurs, the attacker knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the attacker has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The attacker can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded. A version of this attack can be carried out to elevate an external adversary to a full-strength insider adversary as per~\cite{cardestprivacy} (instead of our slightly weaker S4 adversary), since it can simply be done repeatedly each time the external adversary wants to know the complete internals of the HLL sketch. Obvious optimisations can be carried out to reduce the query complexity each time the attack is run (since the bucket contents can only increase over time, some queries are redundant; moreover, the pre-computation is only needed once and is moderate given typical HLL parameters, e.g.\ $n+\ell = 32$).

Since an attacker can escalate from S3 to S4, we focus on scenarios S1, S2 and S4 in what follows. We first present an attack in scenario S1, providing much stronger results than the corresponding attack of~\cite{hllvuln} (in their equivalent scenario M2). We then use the additional information the adversary has access to in scenarios S2 and S4 to make the attack even more powerful.

\subsection{Manipulating the HLL Cardinality Estimate}
\noindent\paragraph{Scenario S1} The goal of the attacker is to infer some information about targeted items' hash values, precisely their bucket mappings, without knowing neither $h$ nor the salt in use (if any). With this knowledge, the adversary is able to retain items that are mapped to a subset of $b$ buckets out of the $2^n$, and minimize the resulting estimated cardinality. To achieve this, the attacker can take advantage of the Hit Counting algorithm initial phase to recover the number of non-empty buckets (whose register is not 0) after inserting few items. Indeed, in the small cardinality range, the cardinality estimator E* is computed as E*$=2^n$ln$(2^n/V)$ with $V$ the number of empty buckets. For typical HLL parameters, e.g.\ $n = 8$, the exact value of $V$, hence $b$, can be recovered from E*. Moreover, when inserting an item in the small cardinality range, the adversary will see a increase in the cardinality estimate if and only if $V$ is modified. This means that the number of empty buckets decreased by 1, therefore that the freshly inserted item is mapped to a bucket that was not hit before.

Thus, if the adversary blindly picks the first $b$ buckets to be hit as its targeted buckets, he can then discard all items whose insertion causes the cardinality estimator to increase from $b+k$ to $b+k+1$ for $k \in \mathbb{N}$. In other words, when $b$ buckets are already hit and the insertion of an item causes the cardinality estimator to increase, the attacker can discard this item as he knows that a new bucket (other than one of the $b$ buckets to be filled) switched to a non-empty state. The adversary can keep inserting items and discarding some while in the small cardinality range.

As soon as the cardinality estimator exceeds $\frac{5}{2}\times 2^n$, the adversary can assume that it corresponds to the raw HLL cardinality estimate with high probability. In this case, no more useful information can be inferred about the bucket mappings, so the attacker has to reset the sketch to the empty state and start again with the non-discarded items. The adversary must pay attention not to shuffle the non-discarded items in between iterations, as the $b$ buckets to be filled are blindly determined by the first inserted items, and must stay the same throughout the attack. After several iterations, the adversary discards all items mapped to unwanted buckets, such that $b$ buckets are filled and the others left untouched.

The value of $b$ has to be picked smartly to keep the final cardinality estimate low, as the registers must hold values potentially very high. $b$ is going to be chosen such that even in the worst case scenario where all register values $M[j]$ hold the the maximal value of $l+1$, the Hit Counting algorithm is going to be used to produce the final estimation. So $b$ must satisfy:
\begin{align}
E:=\alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1} & \le \frac{5}{2}m \nonumber\\
\alpha_mm^2\cdot((m-b)\cdot2^{-0}+b\cdot2^{-(l+1)})^{-1} & \le \frac{5}{2}m \nonumber\\
b & \le m\cdot(1-\frac{2\alpha_m}{5})\cdot(1-2^{-(l+1)})^{-1} \label{eqn:1} 
\end{align}
with $m=2^n$. For example with $n = 8$ and $l = 24$, $b$ can take any value up to 184. The initial set of items to be filtered can be built by varying the flexible fields of the items $x$ the adversary is trying to insert.
%so that the adversary can ``hide'' $2^{t-n}$ elements while keeping the HLL cardinality estimate to 1.

\noindent\paragraph{Scenario S2} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the final averaged approximate cardinality is likely to be left unchanged.
For example, in the scenario of~\cite{portscanhll}, the adversary could change the source port of its IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL sketch without increasing the cardinality markedly, by repeatedly sampling from the input domain.

>>>>

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Kenneth G. Paterson and Mathilde Raynal\\
\vspace{3mm}
Department of Computer Science, ETH Zurich\\
{\tt kenny.paterson@inf.ethz.ch, mraynal@student.ethz.ch}}

@@

\section{Introduction}
Naive approaches to computing the cardinality of large data sets, such as sorting the elements or simply maintaining the set of unique elements seen, are impractical at scale and in streaming settings. The HyperLogLog algorithm (HLL henceforth)~\cite{hll} is today the most widely used cardinality estimator. It is increasingly used in settings where adversaries may have  incentives to manipulate the estimate made by the algorithm. For example,~\cite{portscanhll} proposes its use for detecting network scanning attacks, while Facebook engineers have reported~\cite{fbhll} that they use HLL for estimating the number of active users. These applications demand a careful evaluation of the performance of HLL under adversarial input.

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in the adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of $O(2^t)$ where $t$ is the number of bits of flexibility in the adversary's input).

The rest of the paper is organised as follows. Section~\ref{sec:overview} gives an overview of HLL. A brief summary of  related work can be found in Section \ref{sec:related}. Section~\ref{sec:attacks} presents our different adversarial models. There, we also give attacks and evaluate their impacts under the different adversarial models. The experimental results are discussed in Section~\ref{sec:exp} and Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

@@

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own register, that we call a bucket. We refer to the collection of buckets and their contents as the HLL \emph{sketch}. To map a value $x$ to one of $m=2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$. The output length of $h(\cdot)$ is thus $n+\ell$, fixed to 32 in~\cite{hll}. estimates from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ is empirically computed to correct a systematic multiplicative bias.

Relatively small and big cardinalities are handled separately to correct for systematic distortions, as shown in Fig.~\ref{fig:hll}. When the raw cardinality estimate is below $\frac{5}{2}m$, the Hit Counting algorithm of Whang et al.~\cite{hitcounting} is used to produce the final cardinality estimate, looking at the number of empty buckets (whose register is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hashing collisions. As a result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{hll}.

Some works building on~\cite{hll} replace the Hit Counting algorithm by other schemes and use a hash function with a larger output domain to reduce the chances of collision and thus get rid of the need for large cardinality range correction. For example, \cite{hllpratice} uses a 64-bit hash function and Linear Counting along with bias correction in the small cardinality range. In this report we target the HLL formulation in the original paper~\cite{hll} as represented in Fig.~\ref{fig:hll}, and leave for future work the analysis of its variants.

@@

\begin{figure}[ht]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : \mathcal{D} \rightarrow \{0, 1\}^{32}$ hash data from some domain $\mathcal{D}$ to binary $32$-bit strings.\\
Let $\rho(s)$ be the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^K) = K + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} $\mathcal{M}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $v \in \mathcal{M}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $x := h(v);$\\
    \hspace{2em} \textbf{set} $j = 1 + <x_1x_2...x_{n}>_2$; \hfill \textit{\{the binary address determined by the first $n$ bits of $x$\}}\\
    \hspace{2em} \textbf{set} $w := x_{n+1}x_{n+2}...$;\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the raw HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:=  m\cdot \ln(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range—no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure}

@@

Regardless of the sensitivity of the above examples, none of the above-cited papers are clear about their threat model, especially for the HLL component.  They are all potentially vulnerable to the attacks we present below.

@@

The work of Desfontaines \emph{et al.}~\cite{cardestprivacy} studies the privacy properties of HLL in two different attack scenarios, referred to as \emph{insider} and \emph{external}. The attack target of~\cite{cardestprivacy} is different from ours, being concerned with loss of privacy, whereas we care about manipulating cardinality estimation. It is argued in~\cite{cardestprivacy} that HLL can be used by organisations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm. In Section~\ref{sec:attacks} we will discuss the two attack scenarios from~\cite{cardestprivacy} in more detail. In particular, we will show that an adversary who can insert items into the HLL and who can access the HLL cardinality estimate (via typical APIs in HLL implementations) as permitted in the external scenario of~\cite{cardestprivacy} can actually realise the same attacks as the insider adversary can, at least insofar as cardinality estimate manipulation is concerned.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. We consider the attack model used in~\cite{hllvuln} to be quite artificial, since it gives the adversary the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. After some clarifications obtained through personal correspondence with the authors of~\cite{hllvuln}, we understand that such a setup might be possible under certain circumstances, for example where the adversary has access to an identical ``shadow'' device; see further discussion in Section~\ref{sec:attacks}.

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as a replacement for (but at the cost of losing mergeability) or in addition to (inducing a memory overhead) an unsalted sketch.
%This might not be a convenient solution for some applications, they thus encourage further research.

Clayton \emph{et al.}~\cite{CCS:ClaPatShrPS19} provide a provable-security treatment of probabilistic data structures in adversarial environments more generally. They use their formalism to analyze Bloom filters, counting (Bloom) filters and count-min sketch data structures, but not HLL. It is an interesting open problem to use their framework to develop a formal understanding of how HLL can be made secure in the adversarial setting.

@@

\subsection{Adversarial Scenarios}\label{sec:set}
We model four distinct adversarial scenarios based on the adversary's knowledge and capabilities. To simulate real-life settings, we consider both cases when the sketch is shared and receives inputs from other honest users but also when inputs are under the sole control of the adversary.
It is reasonable to assume that the adversary knows which case it is in. 
These two settings differ mainly in the fact that the adversary is attacking a sketch that may already contain other users' data in the former option, but is empty in the latter.

Among the scenarios, the \textit{insider} setup of~\cite{cardestprivacy} that we present as S4 makes the strongest assumptions about the adversarial capabilities since it assumes that the adversary has a perfect view of the sketch (at some point in time).

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a copy of the HLL instance as a user. It can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The adversary does not know the state of the targeted HLL sketch, meaning that it does not have the list of elements previously inserted into it by other users.
\item[S2:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e.\ it knows the number of buckets $m=2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time. However, it does not know the individual bucket contents.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. Note that, using these capabilities, an S4 adversary can compute a lower bound on the HLL cardinality estimate based on its initial snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

\subsection{Discussion of Adversarial Scenarios}\label{sec:scenarios-discussion}

We do not consider scenario S1 typical of HLL deployments. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the adversary learns which machine is used to do the monitoring, buys the same, and uses it as the shadow device. It is possible that the provider uses a different salt or key in the hash per device. This would make any pre-computation on the items using the shadow device irrelevant, and considerably weaken the attack presented in Section 3.2 of~\cite{hllvuln}. Hence we assume for the rest of this work that, in scenario S1, shadow and targeted devices have the same internals and security parameters. If the HLL device receives inputs from several users, the attack from Section 3.2 of~\cite{hllvuln} is then the easiest to carry out. It is described in~\cite{hllvuln} as follows: "after each insertion [in the shadow device], elements that do not increase the cardinality estimate are retained". The idea is to iterate through the process of trial insertion several times to refine the list of items, before inserting what remains into the targeted device. We explore in this work scenario S1 in the special case where inputs come only from the adversary, and propose an attack that provides better results than the one presented in~\cite{hllvuln}.

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the adversary  inserts a sequence of items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once an increase occurs, the adversary knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the adversary has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The adversary can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded, since then most buckets will store values corresponding to inputs with few leading zero bits. A version of this attack can be carried out to elevate an external adversary to a full-strength insider adversary as per~\cite{cardestprivacy} (instead of our slightly weaker S4 adversary), since it can simply be done each time the external adversary wants to obtain a snapshot of the internals of the HLL sketch. Obvious optimisations can be carried out to reduce the query complexity each time the attack is run (since the bucket contents can only increase over time, some queries are redundant; moreover, the pre-computation is only needed once and is moderate given typical HLL parameters, e.g.\ $n+\ell = 32$).

Since an adversary can quite efficiently escalate from S3 to S4, we focus on scenarios S1, S2 and S4 in what follows. We first present an attack in scenario S1, providing much stronger results than the corresponding attack of~\cite{hllvuln} (in their equivalent scenario M2). We then use the additional information the adversary has access to in scenarios S2 and S4 to make the attack even more powerful.

\subsection{Manipulating the HLL Cardinality Estimate}

Although we consider scenario S1 to be quite artificial, we will commence our analysis there, since the techniques we present are more broadly applicable in S2 and S4.

\noindent\paragraph{Scenario S1} The goal of our adversary is to infer some information about targeted items' hash values, precisely their bucket mappings, knowing neither $h$ nor the salt in use (if any). Using this information, the adversary is able to retain items that are mapped to a subset of $b$ buckets out of the $2^n$, and minimize the resulting estimated cardinality. We assume the adversary either has flexibility in choosing the items to insert, or is presented with a static list of items at the start of its attack. We also assume that the shadow HLL sketch and the target HLL sketch are both initially empty.

Our adversary takes advantage of the use of HLL's small cardinality Hit Counting correction to recover the number of non-empty buckets (whose register does not contain value 0) after inserting a few items into the initially empty shadow HLL sketch. Indeed, in the small cardinality range, the cardinality estimate $E^*$ is computed as $E^*= m \cdot \ln (m/V)$ with $V$ the number of empty buckets. For typical HLL parameters, e.g.\ $n = 8, m =256$, the exact value of $V$, hence $b$, can be recovered from $E^*$. Moreover, when inserting an item in the small cardinality range, the adversary will see an increase in the cardinality estimate for the shadow HLL sketch if and only if $V$ is modified. This in turn means that the number of empty buckets in the shadow HLL sketch decreased by 1, and therefore that the freshly inserted item was mapped to a bucket that was not hit before.

The adversary blindly picks the first $b$ buckets to be hit by simply inserting items into the shadow HLL until the cardinality estimate indicates that $b$ buckets have been touched. It puts these items in a final target list ${\cal L}$. It continues to insert items, discarding all items whose insertion causes the cardinality estimate for the shadow HLL sketch to increase from $b+k$ to $b+k+1$ for $k \in \mathbb{N}$ (here we allow some slack in the total number of buckets that have been hit in order to reduce the number of resets of the shadow device that are required in the attack). In other words, when $b+k$ buckets have already been hit and the insertion of an item causes the cardinality estimate to increase further, the adversary can discard this item as it definitely knows that a new bucket (that is not one of the $b$ targeted buckets) switched to a non-empty state. On the other hand, it adds to the list ${\cal L}$ any item that does \emph{not} cause the cardinality estimate to increase. Such items may have hit one of the initial $b$ buckets, or one of the later buckets; at this stage the adversary cannot be certain. The adversary can keep inserting items and discarding those which cause the cardinality estimate to increase while in the small cardinality range.

As soon as the cardinality estimate exceeds $\frac{5}{2}m$, the adversary can assume that the shadow HLL sketch's output corresponds to the raw HLL cardinality estimate with high probability. In this case, no more useful information can be inferred about the bucket mappings in the shadow HLL sketch, so the adversary resets the shadow HLL sketch to the empty state and starts again with the list ${\cal L}$ of non-discarded items obtained so far. Reinserting these items will (eventually) determine whether they hit the original $b$ buckets or later buckets, and hence whether they should be retained in ${\cal L}$ or discarded. Once ${\cal L}$ is reprocessed (and assuming the cardinality estimate does not exceed $\frac{5}{2}m$), the adversary can start to insert fresh items, as previously described. The adversary must be careful not to shuffle the non-discarded items in between iterations, as the $b$ buckets to be filled are blindly determined by the first inserted items, and must stay the same throughout the attack.

After sufficiently many iterations of this procedure, the adversary will have created a list of items ${\cal L}$ such that, when inserted into the \emph{target} HLL, create a sketch in which $b$ target buckets are filled and the others left empty. With an appropriate choice of $b$ (as discussed immediately below), the final cardinality estimate for the target HLL sketch will be $E^*= m \cdot \ln (m/V)$ where $V = m-b$ is the number of empty buckets. This can be rewritten as $E^*= m \cdot \ln (1+b/(m-b))$.
